---
title: "Advanced Quantitative Analysis (BF998) Report Enwushe, Kelvin Buque (202351865)"
output:
  pdf_document: default
  html_notebook: default
  html_document: default
  word_document: default
---
```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

#Setting up Working Tools 
Installing Relevant Packages
```{r}
# install.packages("haven")    # for loading sav file
# install.packages("mice")    # for multiple imputation
# install.packages("lmtest", dependencies = TRUE) # for systemfit
# install.packages("sandwich") # for robust variance estimation in linear regression models.
# install.packages("olsrr") # for Breusch Pagan Test for Heteroskedasticity
# install.packages("car") # for lavene's Test of equality of variance
# install.packages("dplyr") # for data manipulation (mutation) - adding new variable
# install.packages("naniar")  # for MCAR little test
# install.packages("e1071") for skewness
# install.packages("rcompanion") # for overlaying histogram with normal distribution curve
```

## Setting up Working Directory
```{r}
setwd("C:/Users/jjb23183/OneDrive - University of Strathclyde/Documents/PHD COURSES (RM)/BF998 - ADVANCED/BF998 - R File")
```

## Loading Libraries
```{r}
# Loading Libraries 
library(haven)
library(mice)
library(lmtest)
library(sandwich)
library(olsrr)
library(car)
library(dplyr)
library(VIM) # pattern of missingness
library(broom)
library(naniar) 
library(e1071)
library(rcompanion)
```


# Analysis
## Pre-Analysis (Data Exploration)
Loading and Preview of Original Dataset
```{r}
data <- read_sav("Assigment Dataset.sav")

dim(data) #Total number of observations and variables in the dataset

nrow(na.omit(data)) # Number of rows with complete values
```

Extraction Relevant Variables from Original Dataset
```{r}
extracted_data <- data[c("a8a", "a8b", "a8c", "a8d", "a8e", "a8f", "a8g", "b8a", 
                         "b8b", "b8c", "c1a", "c1b", "c1c", "e9_13", "e1", "e2")]

dim(extracted_data) # number of rows and variables in new dataset
summary(extracted_data)
nrow(na.omit(extracted_data)) # For number of rows with complete values
sum(is.na(extracted_data)) # total count of missing values in the entire dataset
nrow(na.omit(extracted_data)) / nrow(extracted_data) # Proportion of Complete Observation

```
Calculating MCAR little test and Exploring the Pattern of Missingness
```{r}
# Perform Little's MCAR test
mcar_test_result <- mcar_test(extracted_data)
mcar_test_result

# Distribution of Missing Rows
#md.pattern(extracted_data)

# Pattern of Missing Values
aggr(extracted_data, col=c("green", "red"))
```

Creating New Variables - Mean of Explanatory Variables in the Dataset
```{r}
extracted_data <- mutate(extracted_data,
                         mean_Jobsat = rowMeans(select(extracted_data,
                         "a8a", "a8b", "a8c", "a8d", "a8e", "a8f", "a8g")))
extracted_data <- mutate(extracted_data,
                         mean_Manview = rowMeans(select(extracted_data,
                         "b8a", "b8b", "b8c")))
extracted_data <- mutate(extracted_data, 
                         mean_Workersview = rowMeans(select(extracted_data,
                        "c1a", "c1b", "c1c")))

# Creating a new dataframe of Relevant Variables
#write_sav(extracted_data, "extracted_file.sav")
```

 
Data Cleaning (Using Multiple Imputation with Mice Package)
```{r}
imputed_data <- mice(data=extracted_data, m = 20, maxit = 10, seed = 20000,
                print = FALSE)
summary(imputed_data)
imp.dataset <- complete(imputed_data, "long") # List of the imputed dataset
```
Where: 
m = number of imputations
maxit  = maximum number of iterations for each imputed dataset
seed = reproducibilty of the imputation(that is, if set at 20000, it will produce the same imputed dataset).

Checking the Number of Complete Values in Imputed Dataset
```{r}
nrow(imp.dataset) # Number of Observations
nrow(na.omit(imp.dataset)) # Number of Completed Observations
```

## Analysis
3.3 DESCRIPTIVE STATISTICS OF OUTCOME VARIABLE (JOB SATISFACTION)
```{r}
# Descriptive statistics of all dataset
summary(imp.dataset)
```

Descriptive Statistics of Job Satisfaction (Outcome Variable)
```{r}
# Summary Statistics of Job Satisfaction
summary(imp.dataset$mean_Jobsat)
sd(imp.dataset$mean_Jobsat)
skewness(imp.dataset$mean_Jobsat)

# Plotting Histogram with Overlay - Dependent Variable
plotNormalHistogram(imp.dataset$mean_Jobsat, prob = FALSE, 
                    main = "Job Satisfaction Histogram", 
                    xlab = "Job Satisfaction", linecol="green")


# Summary Statistics of Job Satisfaction by Gender (Sex)
summary(imp.dataset$mean_Jobsat[imp.dataset$e1==1]) # Male
sd(imp.dataset$mean_Jobsat[imp.dataset$e1==1])

summary(imp.dataset$mean_Jobsat[imp.dataset$e1==2]) # Female
sd(imp.dataset$mean_Jobsat[imp.dataset$e1==2])


boxplot(imp.dataset$mean_Jobsat ~ imp.dataset$e1,
  main = "Job Satisfaction Distribution by Gender (Sex)",
  xlab = "Gender (Sex)",
  ylab = "Mean",
  col = "green",
  border = "black",
  names = c("Male", "Female"))


# Summary Statistics of Job Satisfaction by Age
summary(imp.dataset$mean_Jobsat[imp.dataset$e2])
sd(imp.dataset$mean_Jobsat[imp.dataset$e2])
skewness(imp.dataset$mean_Jobsat[imp.dataset$e2])

# Summary of different ages
summary(imp.dataset$mean_Jobsat[imp.dataset$e2 == 1]) 
summary(imp.dataset$mean_Jobsat[imp.dataset$e2 == 2])
summary(imp.dataset$mean_Jobsat[imp.dataset$e2 == 3])
summary(imp.dataset$mean_Jobsat[imp.dataset$e2 == 4])
summary(imp.dataset$mean_Jobsat[imp.dataset$e2 == 5])
summary(imp.dataset$mean_Jobsat[imp.dataset$e2 == 6])
summary(imp.dataset$mean_Jobsat[imp.dataset$e2 == 7]) 
summary(imp.dataset$mean_Jobsat[imp.dataset$e2 == 8])
summary(imp.dataset$mean_Jobsat[imp.dataset$e2 == 9])

boxplot(imp.dataset$mean_Jobsat ~ imp.dataset$e2,
  main = "Job Satisfaction Distribution by Age",
  xlab = "Age",
  ylab = "Mean",
  col = "green",
  border = "black",
  names = c("16-17", "18-19", "20-21", "22-29", "30-39", "40-49",
          "50-59", "60-64", "65 or More"))
```



INDEPENDENT SAMPLES T-TEST
```{r}
boxplot(imp.dataset$mean_Jobsat ~ imp.dataset$e9_13,
  main = "Job Satisfaction Distrubtion by Computer Use",
  xlab = "Computer Use",
  ylab = "Mean Job Satisfaction",
  col = "green",
  border = "black",
  names = c("Use Computer at Work", "Dont Use Computer at Work"))

# Convert e9_13 to a factor variable
imp.dataset$e9_13 <- factor(imp.dataset$e9_13) # to test the equality of variance

# Levene's Test for equality of variances
leveneTest(mean_Jobsat ~ e9_13, data = imp.dataset)

# Perform Welch's t-test
t.test (mean_Jobsat ~ e9_13, data = imp.dataset)

```

REGRESSION
```{r}
# Calculating Correlation between Outcome and explanatory variables
cor(imp.dataset$mean_Jobsat, imp.dataset$mean_Manview)
cor(imp.dataset$mean_Jobsat, imp.dataset$mean_Workersview)
```

Regression Analysis
```{r}
# Before multiple Imputation
extracted_data.reg <- lm(mean_Jobsat ~ mean_Manview + 
                      mean_Workersview, data=extracted_data)
summary(extracted_data.reg)

# Imputed Data
reg.model.mi <- with(imputed_data, lm(mean_Jobsat ~ mean_Manview + mean_Workersview))
summary(reg.model.mi) # Run Code to view all 20 imputations estimates

summary(reg.model.mi$analyses[[1]]) # Regression for 1st imputation - change the 
                                  # number in the bracket to view various imputations

# Adding Control 
reg.model.mi2 <- with(imputed_data, lm(mean_Jobsat ~ mean_Manview +
                                    mean_Workersview + e1)) # Gender
summary(reg.model.mi2)

reg.model.mi3 <- with(imputed_data, lm(mean_Jobsat ~ mean_Manview +
                                    mean_Workersview + e2)) # Age 
summary(reg.model.mi3)

reg.model.mi4 <- with(imputed_data, lm(mean_Jobsat ~ mean_Manview +
                                    mean_Workersview + e1 + e2)) # Gender + Age
summary(reg.model.mi4)
```
* Run the block of codes above to view detail estimates of all 20 imputations 
for each model. 


Pooling Regression Estimate
```{r}

pooled.fit <-  pool(reg.model.mi)
pooled.fit2 <- pool(reg.model.mi2)
pooled.fit3 <- pool(reg.model.mi3)
pooled.fit4 <- pool(reg.model.mi4)

summary(pooled.fit)
summary(pooled.fit2)
summary(pooled.fit3)
summary(pooled.fit4)
```

Computing pooled r-square 
```{r}
# Computing r-Squared
# Extract model statistics from each imputed dataset

pool.r.squared(pooled.fit)
pool.r.squared(pooled.fit2)
pool.r.squared(pooled.fit3)
pool.r.squared(pooled.fit4)
```

## Robustness Estimation Testing for Best FIt Model
Note: To test assumptions, we ran each of the 20 imputations in the selected model 
(reg.model.mi) separately to generate the average for VIF and multicollinearity
test. Also in the test for normality and linearity, we also created several plot by changing the number of imputations to examine the distribution of the fitted and residual values.

The purpose was because the average of these imputations cannot be generated automatically.The code below represents the first regression imputations out of 
the 20 in the selected model. To view other imputations, like say the fifth, all we need to do is change the value in the bracket to 5 or any other number. 

```{r}
# Test 1 - Linearity and Normality
par(mfrow = c(2, 2))
# Use plot() function to create diagnostic plots
plot((reg.model.mi$analyses[[1]]))

hist(reg.model.mi$analyses[[1]]$residuals) # Normality
```



```{r}
# Test 2 - Breusch Pagan Test for Heteroskedasticity
ols_test_breusch_pagan(reg.model.mi$analyses[[1]])
```


```{r}
# Test 3- Variance Inflation Factor (VIF) for each independent variable  - For
# Multicollinearity

car::vif(extracted_data.reg) # Extracted data before MI

car::vif(reg.model.mi$analyses[[1]])  # Extracted data after MI 
```
The code above represents the first regression imputations out of the 20 in the
selected model. TO view other imputations, like say the fifth, all we need to do
is to change the value in the bracket to 5. 
The average VIF had to be computed manually because it could not be automatically generated from pooled regression models.






