---
title: "BF994 Assessment: Enwushe, Kelvin Buque - 202351865"
output:
  word_document: default
  html_notebook: default
  html_document: default
  pdf_document: default
---
```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```
#######################################################              
#     PART 1 - INTRODUCTION/DESCRIPT.IVE STATISTICS    #
#######################################################
# The Determinants of Job Satisfaction
## 1.0 Background
                        
This analysis is about understanding the factors that influence job satisfaction.The
literature on employee job satisfaction is robust as scholars have provided different 
perspectives on the debate (Cornelißen, 2006). This is particularly true because 
employees (humans) are considered the most important resource in organisations. Meier 
and Spector (2015, p. 1) define job satisfaction “as a person’s overall evaluation of 
his or her job as favourable or unfavourable”. Scholars have argued and established 
link between employee satisfaction and firm performance (Meier and Spector (2015; 
Sainju, Hartwell and Edwards, 2021). Furthermore, unsatisfied employees lead to a high
attrition rate (Pablo, 2023), and the cost of replacement often puts a big burden on 
an organisation operationally and financially (Kim et al., 2023).

Employee attrition in the workplace is at an all-time high. One possible explanation 
for this could be the technology disintermediation and the rise of the “gig economy” facilitated by the COVID-19 pandemic. As a result, employees have more options than 
ever before. This has become a challenge for managers. It is therefore imperative to understand the various important factors that can influence job satisfaction to help
managers of organisations understand the intrinsic and extrinsic nature of their 
employees and develop strategies to facilitate retention. Hence the questions: What 
are the determinants of job satisfaction? Furthermore, most studies on job satisfaction 
have been done using other traditional approaches (ANOVA, Structural Equation Modelling, Qualitative Comparative Analysis Using Fuzzy Sets (FsQCA)), with few studies using
econometric methods.


## Dataset

Using a teaching dataset from an abridged version of a 2004 Workplace Employment
Relations Survey by Stokes and Forth (2008), accessed from the UK Data Service Archive,
the question (What are the determinants of job satisfaction?) will be addressed.
The contents of the dataset are responses from workplaces and employees collected
through a cross-sectional survey. It contains around 400 variables (columns) and
22451 cases (observations).

## Measures

==> Dependent variables
a8g - How satisfied are you with the work itself?

a8g is already given as the dependent variable in this assessment. Hence the outcome variable.

==> Independent variables (explanatory).
From the literature, several variables were identified with potential to influence job satisfaction: job security (Cornelißen, 2006; Kim et al., 2023), autonomy (Cornelißen, 2006) and employer-employee relationship (Cornelißen, 2006; Alegre, Mas-Machuca and Berbegal-Miraben, 2016). These articles were chosen because they capture studies on job satisfaction determinants in both private and public sector. 

For this study, relationship (employer-employee), job security, and autonomy (using initiative) were used as the main explanatory variables. These variables were selected because they align with the two-factor theory developed by Friedrick Herzberg, which has been widely deployed by researchers in understanding job satisfaction determinants (Lee et al., 2023). 

a8b = Using initiative (Autonomy)
a8f = Job security
c3 = Relationship

## 1.1 Packages, Libraries and Dataset

```{r}
######################################
# Install All relevant Packages      #
######################################
# install.packages("haven")    # for loading dta file
# install.packages("moments")  # for measuring Skewness and Kurtosis
# install.packages("rcompanion") # for overlaying histogram with normal distribution curve
# install.packages("lmtest", dependencies = TRUE) # for systemfit
# install.packages("sandwich") # for robust variance estimation in linear regression models.
# install.packages("AER", dependencies = TRUE)   # for running IV models
# install.packages("systemfit", dependencies = TRUE)     # needed for running IV models
# install.packages("stargazer") # for better presentation of result
```

```{r}

# Setting up Working Directory
setwd("C:/Users/jjb23183/Downloads/ASSESSMENT DATASET")
```

```{r}
# Loading Libraries 
library(haven)
library(moments)
library(rcompanion)
library(lmtest)
library(sandwich)
library(systemfit)
library(AER)
library(stargazer)
```

```{r}
# Loading Dataset
employment_survey <- read_dta("xs04_seqv2_teaching_dataset.dta")
```

## 1.2 Cleaning Dataset
```{r}
employment_survey_clean <- subset(employment_survey, a8g >= 1)
employment_survey_other <- subset(employment_survey, a8g <= -1) # Not answered (-9) + Don't know (-8) + Multi-coded (-6) +  Item not applicable (-1)
nrow(employment_survey_clean) / nrow(employment_survey) * 100
```
After pre-processing the dataset, the clean data set represents about 99% of the observations with only 1% for missing data and otherwise. The purpose of this process was to improve the improve the efficiency of the analysis and limit estimation bias.

## 1.3 Performing Descriptive Statistics on Job Satisfaction (a8g)

```{r}
summary(employment_survey$a8g)
sd(employment_survey$a8g)
var(employment_survey$a8g)
skewness(employment_survey$a8g) 
kurtosis(employment_survey$a8g)

# Plotting Histogram with Overlay
plotNormalHistogram(employment_survey$a8g, prob = FALSE, 
                    main = "Job Satisfaction Histogram with Overlay", 
                    xlab = "Job Satisfaction", linecol="green")

# Plotting the dependent variable with the selected independent variables
plot(employment_survey$a8f,employment_survey$a8g, xlab="Job Security", ylab="Job Satisfaction", col="green") # Job security
plot(employment_survey$a8b,employment_survey$a8g, xlab="Using Initiative (Autonomy)", ylab="Job Satisfaction", col="green") # Using Initiative (Autonomy)
plot(employment_survey$c3,employment_survey$a8g, xlab="Relationship", ylab="Job Satisfaction", col="green") # Relationship
```
From the sample dataset, median value is 2, indicating that a significant number of employees are satisfied with their job. The distribution is also skewed to the left with a negative value of -4.32.  


# Distribution by gender and Age
```{r}
# Job Satisfaction by Gender
length(employment_survey$a8g) # Variable length

job_satisfaction_male <- length(employment_survey$a8g[employment_survey$e1 == 1])
job_satisfaction_male 

job_satisfaction_female <- length(employment_survey$a8g[employment_survey$e1 == 2])
job_satisfaction_female

job_satisfaction_notAnswered <- length(employment_survey$a8g[employment_survey$e1 == -9])
job_satisfaction_notAnswered 


percentage_male <- job_satisfaction_male / length(employment_survey$a8g_length)
percentage_male

percentage_female <- job_satisfaction_female / length(employment_survey$a8g_length)
percentage_female

percentage_notAnswered  <- job_satisfaction_notAnswered  / length(employment_survey$a8g_length)
percentage_notAnswered 

# Mean distribution 
tapply(employment_survey$a8g, employment_survey$e1, summary) 
tapply(employment_survey$a8g, employment_survey$e2, summary) 
```

The result shows that that female respondents represented 53.3% of the votes, men 46.2% and the rest unanswered.  It also shows the median job satisfaction for male and female is equal. However, the mean for male is slightly higher than that of female.

In the case of Age, the mean job satisfaction is higher for employees with ages 22-29  (4), followed by 18-19 (2) and lowest for employees at 65 and above (9). This is not surprising. Possible explanation for this could be that workers at 22-26 are relatively fresh in the workforce and thus would be more interested in building a career.However, the latter are close to retirement and most likely, diminished excitement.


##########################################                      
#     PART 2 - REGRESSION ANALYSIS       #
##########################################

## 2.1 Calculating Correlation
```{r}
cor(employment_survey$a8g, employment_survey$a8f)
cor(employment_survey$a8g, employment_survey$a8b)
cor(employment_survey$a8g, employment_survey$c3)
```
The correlation coefficient for job satisfaction and job security is 0.44, job satisfaction and using initiative (autonomy) = 0.58, and job satisfaction and relationship = 0.2 respectively. While the coefficients does not suggest a strong relationship (employer-employee) between the outcome and explanatory variables, it is however, not uncommon in regression analysis. Uncorrelated variables can solve the challenge of duplication. 

## 2.2 Regression of Job Satisfaction on job security, using initiative (autonomy) and relationship (employer-employee)

```{r}
# First Regressing Job Satisfaction on using initiative (autonomy)
lin_reg <- lm(a8g ~ a8b, data=employment_survey)
summary(lin_reg)

# Adding one variable
mul_regression1 <- lm(a8g ~ a8b + a8f,  data=employment_survey) # using initiative (autonomy) + job security 
summary(mul_regression1)

# Adding 2 more variables
mul_regression2 <- lm(a8g ~ a8b + a8f + c3,  data=employment_survey) # using initiative (autonomy) + job security + relationship (employer-employee)
summary(mul_regression2)

stargazer(lin_reg, mul_regression1, mul_regression2, title=" Regression Results", type = 'text')


stargazer(lin_reg, mul_regression1, mul_regression2, title=" Regression Models", type = 'text', dep.var.labels=c("Job Satisfaction"), covariate.labels=c("using initiative (autonomy)",
"job security ","Relationship (employer-employee"))

```


## 2.2.1 Extracting R-Squared Values from Models and Calculating Computing with Akaike
information  criterion (AIC)
```{r}
# R-Squared
summary(lin_reg)$r.squared
summary(mul_regression1)$adj.r.squared
summary(mul_regression2)$adj.r.squared

# Computing with Akaike information criterion
AIC(lin_reg)
AIC(mul_regression1)
AIC(mul_regression2)
```

Following regression analysis, results of all three models suggest statistical significance between the dependent and explanatory variables with p-values less than 0.05. Additionally, as more explanatory variables were added, the r-square continually increased, albeit slightly. That is, the values increased from 0.33 in Model 1 (lin_reg) to 0.38 in Model 2 (mul_regression1) and 0.39 in Model 3 (mul_regression2). However,the difference in Model 2 and Model 3 is negligible as they both fit the data.
To explicate and choose the right model, Akaike Information Criterion (AIC) was computed to know if adding another explanatory variable in Model 3 was necessary because of the negligible increase in r-square. From the AIC test, it shows Model 3 has a lower AIC score, meaning that the additional variable added in Model 3 improved the model. Hence signifying a better fit.


## 2.3 Running Robust Estimation Tests

2.3.1 Measuring F- Test (Evauate the significance of the overall model)

* Estimating full Model (mul_regression2) ....See 2.2

* Estimating Restricted Model (mul_regression2) ....See 2.2

```{r}
anova(mul_regression2, lin_reg)
```

2.3.2 Running RESET  Test to (Evaluate Model Adequacy)

```{r}
resettest(mul_regression2)
```

2.3.3 Performing Robustness Test - Estimating Heteroskedasticity - Robust Standard Error 

```{r}
# creating residual vs. fitted plot to check for Heteroskedasticity
par(mfrow=c(2,2)) # init 4 charts in 1 panel
plot(mul_regression2,  col = "green")

# Testing Heteroskedasticity significance using ols_test_breusch_pagan Test
# install.packages("olsrr")
# library(olsrr)
ols_test_breusch_pagan(mul_regression2)


# Calculating Robust Standard Errors (addressing the Heteroskedasticity issues)
coeftest(mul_regression2, vcov=vcovHC(mul_regression2, "HC1"))

stargazer(mul_regression2, coeftest(mul_regression2, vcov=vcovHC), title="Robust Standard Errors for Models", type = 'text', dep.var.labels=c("Job Satisfaction"), covariate.labels=c("using initiative (autonomy)", "job security ","Relationship (employer-employee"))

```

Having earlier established that the coefficients of all the explanatory variables in the model are significant in explaining Job satisfaction from the p-values (2e-16 ~ 0), the p-value from the F-test
also suggests statistical significance. Meaning that, the null  hypothesis should be rejected with a level of confidence, and the Model (mul_regression2) be accepted.

The output of the p-value test (2e-16) suggest a strong evidence against the null hypothesis which  implies that there are no overlooked nonlinearities within the model. However, the low p-value might  suggest that the model might not fully capture the relationship between the variables.
.
TThe ols_test_breusch_pagan test suggest the presence of heteroskedasticity - the dispersion of variance of the residuals fluctuates across distinct levels of the independent variables. The p-value of 0 in this test signifies substantial evidence against the null hypothesis.- That is, it indicates the presence of heteroscedasticity within the residuals. 

From the robustness test, the standard errors generated are more than the standard errors of the original model suggesting that the latter was underestimated as a consequence of the heteroskedasticity.The former represents the most reliable estimates of the standard errors of the regression coefficients.
 

##########################################                      
#     PART 3 - CAUSAL INFERENCE          #
##########################################
This should include an explanation and discussion of omitted variable bias, and conditioning on observables.

```{r}
mul_regression2 <- lm(a8g ~ a8b + a8f + c3,  data=employment_survey) # using initiative (autonomy) + job security + relationship (employer-employee)
summary(mul_regression2)
```


mitted variable bias could arise if there are other important factors influencing job satisfaction that are not included in the analysis. These factors could correlate with one or more of the independent variables and also affect job satisfaction, leading to biased estimates of their effects.









##########################################                      
#   Bibliography                        #
##########################################

Alegre, F., Mas-Machuca, M. and Berbegal-Miraben, J. (2016) “Antecedents of Employee Job Satisfaction: Do they Matter?”, Journal of Business Research, 69, pp. 1390-1395.

Cornelißen, T. (2006) “Job Characteristics as Determinants of Job Satisfaction and Labour Mobility” Diskussionsbeitrag, No. 334, Universität Hannover, Wirtschaftswissenschaftliche Fakultät, Hannover. Available at: http://hdl.handle.net/10419/22446 (Accessed: 3 March 2024).

Hlavac, M. (2022) “Stargazer: Well-Formatted Regression and Summary Statistics Tables”, R package version 5.2.3. https://CRAN.R-project.org/package=stargazer

Kim, L., Pimlapas, P., Siwarit, P., Ngachonpam, H. and Vikas, K. (2023) “Key Determinants of Job Satisfaction among University Lecturers”, Social Sciences, 12, 153, pp. 1-16. https://doi.org/ 10.3390/socsci12030153.

Lee et al. (2023) “Exploring Determinants of Job Satisfaction: A Comparison Between Survey and Review Data”, SAGE Open. https://doi.org/10.1177/21582440231216528

Meier, L.L. and Spector, P.E. (2015) “Job Satisfaction. In Wiley Encyclopedia of Management (eds C.L. Cooper, D.E. Guest and D.J. Needle). https://doi.org/10.1002/9781118785317.weom050093.

Pablo, J. M. (2023) “Continuing Professional Development: An Overlooked Determinant of Job Satisfaction”, International Journal of Management and Development Studies, 12(3), pp. 01-15. DOI: 10.53983/ijmds.v12n03.001

Sainju, B., Hartwell, C. and Edwards, J. (2021) “Job Satisfaction and Employee Turnover Determinants in Fortune 50 Companies: Insights from Employee Reviews from Indeed.Com”, Decision Support Systems, 148, 113582. https://doi.org/10.1016/j.dss.2021.113582.

Stokes, L. and Forth, J. (2008) Workplace Employment Relations Survey, 2004: Teaching Dataset, [data collection], Department of Trade and Industry, Employment Market Analysis and Research, 2nd Edition, Department of Trade and Industry, Employment Market Analysis and Research, [original data producer(s)]. Available at: DOI: http://doi.org/10.5255/UKDA-SN-5897-1 (Accessed: 3 March 2024).


